\textbf{Differences in COVID-19 vaccination in the province of Ontario across Health Regions and socio-economic strata}

### 2023-05-12 manuscript_v5/01-background.qmd
updated, working on discussion

This is particularly important in the case of Ontario, the most populated province in Canada. Between 2006 and 2019, Ontario provided healthcare access to its inhabitants using 14 intra-provincial divisions called the Local Health Integrated Networks (LHINs). However, due to is complexity, bureaucratic burden, and lack of collaborative work this approach resulted in disparities such as excessive expenditures and differences in mortality rates [@tsasis2012;@muratov2018]. In an effort to improve the healthcare system of the province, in late 2019 the 14 LHINs were phased out and the areas they covered were incorporated into 6 Health Regions (North East, North West, Central, Toronto, West, and East)  [@dong2022]. Due to the relatively recent adoption of the Health Region model and its alignment with the onset of the COVID-19 pandemic, there is a need to analyze if there are ongoing disparities in health access under this approach that need to be addressed before they are exploited by a new disease or public health threat. In this regard, analyzing differences in vaccination uptake within the Health Regions and can help identify which socio-demographic groups are the most vulnerable and what areas of the province deserve special attention by decision-makers.

>>>>

Between 2007 and 2019, Ontario provided healthcare access to its inhabitants using 14 intra-provincial divisions called the Local Health Integrated Networks (LHINs). However, this approach was complex and bureaucratic, resulting in excessive expenditures, disparities in mortality rates, and the deterioration of certain performance indicators such as wait times and readmissions [@tsasis2012;@muratov2018;@dong2022]. In an effort to improve the healthcare system of the province, in late 2019 the government dissolved the 14 LHINs  and the areas they covered were incorporated into 6 Health Regions (North East, North West, Central, Toronto, West, and East)  [@dong2022]. Due to the relatively recent adoption of the Health Region model and its alignment with the onset of the COVID-19 pandemic, there is a need to analyze if there are ongoing disparities in health access under this approach that need to be addressed before they are exploited by a new disease or public health threat. In this regard, previous research has highlighted disparities in the level of activity of each Health Region [@sethuram2023]. Therefore, analyzing differences in vaccination uptake within the Health Regions and can help identify which socio-demographic groups are the most vulnerable and what areas of the province deserve special attention by decision-makers.

### 2023-05-12 manuscript_v5/03-results.qmd
updated, working on discussion

@@

@fig-model-uncorr presents the estimates (as odd ratios) from the logistic regression models for vaccination status using the socio-demographic factors collected by the survey, and their interactions. Generally speaking, lower odds of vaccination were identified in both cases in individuals characterized by a low household income, or that identified as part of equity-deserving groups. However, the magnitude of the estimates differed between the uncorrected and corrected models and more importantly, certain estimates were not deemed to be statistically-significant after the correction, in contrast to the estimates from the uncorrected model. Specifically, the corrected model showed no significant differences in vaccination odds between the age groups considered, the East Health Region, Latin American individuals with a household income under CAD 25,000, and Indigenous individuals living in the Central Health Region (@fig-model-uncorr,B).
<!-- \newpage -->

@@

```{r,models}
#| message: false
#| warning: false
#| echo: false
#| fig-align: left
#| fig-width: 15
#| fig-height: 12
#| label: fig-model-uncorr
#| fig-cap: "Coefficient estimates and confidence intervals for the uncorrected model. Only statistically significant interaction terms are shown. Full interaction terms can be found in Supplementary Figure A-3."

@@

#print(fig_mod, vp=viewport(angle=90,height = 0.5,clip="off"))
fig_mod

>>>>

# test<-a1_rake$variables
# 
# a1_rake %>%
# tbl_svysummary(percent="row",
#               by= first_dose,
#               include=c(income_ord,
#                         Age_group_ord,
#                         Health_Region,
#                         Month,
#                         Race),
#               label=list(Health_Region ~ "Health Region",
#                  Age_group_ord ~ "Age Group",
#                  income_ord="Income (CAD)")
#   ) %>% add_p(test = everything () ~ "chisq.test")%>%
#   bold_labels() %>%
#   modify_table_styling(
#     columns=label,
#     rows=label=="Other",
#     footnote="Southeast Asian, Filipino, West Asian, \nand minorities not identified elsewhere according to the Census."
#   ) %>%
#   modify_header(label = "**Variable**") %>%
#   as_gt()
#

@@

@fig-models presents the estimates (as odd ratios) from the logistic regression models for vaccination status using the socio-demographic factors collected by the survey, and their interactions. Generally speaking, lower odds of vaccination were identified in both cases in individuals characterized by a low household income, or that identified as part of equity-deserving groups. However, the magnitude of the estimates differed between the uncorrected and corrected models and more importantly, certain estimates were not deemed to be statistically-significant after the correction, in contrast to the estimates from the uncorrected model. Specifically, the corrected model showed no significant differences in vaccination odds between the age groups considered, the East Health Region, Latin American individuals with a household income under CAD 25,000, and Indigenous individuals living in the Central Health Region (@fig-models,B).
<!-- \newpage -->

@@

```{r,models}
#| message: false
#| warning: false
#| echo: false
#| fig-align: left
#| fig-width: 15
#| fig-height: 12
#| label: fig-models
#| fig-cap: "Coefficient estimates and confidence intervals for the uncorrected model. Only statistically significant interaction terms are shown. Full interaction terms can be found in Supplementary Figure A-3."

@@

fig_mod

#table for comparison
#tbl_merge(tbls = list(tbl_regression(m0,exponentiate=TRUE),tbl_regression(m2,exponentiate = TRUE)))
```

### 2023-05-12 manuscript_v5/04-discussion.qmd
updated, working on discussion

The existence of healthcare disparities in Ontario motivated the recent change in the healthcare system of the province, which switched from the LHIN model to the Health Region model in late 2019  [@tsasis2012;@dong2022]. This change is relatively recent, and because of this, is likely that there are ongoing differences in healthcare access across the province. In this study we hypothesized that differences in COVID-19 vaccination uptake were present between the Health Regions, aiming at determining which socio-demographic groups could be impacted by these disparities and to provide decision-makers with information that could be used to develop policies focused on reducing or eliminating these differences and ensure that the Health Region model is able to improve health access for the inhabitants of Ontario, a province that faces unique challenges due to its condition as the most populated and the most ethnically diverse province of Canada.

Our results indicate that across the most densely populated Health Regions of Ontario, almost three quarters of surveyed individuals reported to have received the first dose of the COVID-19 vaccine (@tbl-descriptive-stats). We identified significant intra-provincial differences in vaccination based on socio-economic and geographical factors. First, our results show differences in odds of vaccination in individuals with a household income below CAD 60,000 and in individuals belonging to visible minority groups. Those who identified as Arab/Middle Eastern, Black, Latin American, or  that belonged to a minority group not included in the survey (Southeast Asian, Filipino, West Asian, and minority groups not identified elsewhere) had vaccination odds that were less than a third of individuals that identified as White/Caucasian (@fig-model-corr). These results are consistent with other studies that have shown lower vaccination rates in individuals that identify as part of a racial minority, or that have a low household income [@guay2022;@muhajarine2021;@carter2022;@hussain2022]. In the case of Ontario, a possible rationale for this difference is vaccine access, which has been identified as an important contributing factor of disparities in vaccination for equity-deserving groups [@njoku2021], and that occurred in Toronto earlier in 2021, where it has been shown that vaccination venues were scarce in low socio-economic areas that had the highest burden of COVID-19 at the time [@bogoch2022].

>>>>

Between 2007 and 2019, healthcare in Ontario was managed by 14 different LHINs, which aimed to provide an integrated health system for the province. However, this approach faced chronic problems such as fragmented electronic health systems [@lysyk2016], the decline of performance indicators (e.g., increased unplanned emergency visits, number of patient readmissions), and inequities in health services access [@lysyk2015]. Based on these problematics and with the intent of better organizing and delivering care, the provincial government moved to the Health Region model in late 2019 [@dong2022].

In this study we hypothesized that differences in COVID-19 vaccination uptake were present between the Health Regions, aiming at determining which socio-demographic groups could be impacted by these disparities in order to provide decision-makers with information that could be used to develop policies focused on reducing or eliminating these differences and ensure that the Health Region model is able to fulfill its missing of improving health access for the inhabitants of Ontario.

Our results show differences in odds of vaccination in individuals across certain socio-demographic groups. Specifically, those who identified as Arab/Middle Eastern, Black, Latin American, having mixed racial or ethnic background, or that belonged to other groups not explicitly included in the survey (Southeast Asian, Filipino, West Asian, and minority groups not identified elsewhere) had vaccination odds that were between a third and a half of that of individuals that identified as White/Caucasian (@fig-models). These results are consistent with previous studies that have shown lower vaccination rates in individuals with the same socio-demographic characteristics [@guay2022;@muhajarine2021;@carter2022;@hussain2022].

Lower vaccine uptake in the socio-demographic groups indicated above may be influenced in part, by vaccine hesitancy and refusal [@cnat2022a;@cnat2022b], which have been associated in Canadian individuals to concerns on vaccine safety and effectiveness [@basta2022]. However, it is also likely that accessibility played an important role in vaccine uptake, as it has been shown that this issue plays an important role in the case of equity-deserving groups [@njoku2021]. Although a spatial analysis of vaccine accessibility in Ontario is beyond the scope of this study, it is interesting to note that vaccination venues were scarce in low socio-economic areas that had the highest burden of COVID-19 at the time in Toronto and other regions of Ontario [@bogoch2022;@iveniuk2021].

### 2023-05-12 manuscript_v5/appendix.qmd
updated, working on discussion

Tables A-7 and A-8 present summaries of information from the Fields survey (from the clean dataset analyzed in the main paper), showing responses by White/Caucasian individuals by Health Region and proportion of answers by LHIN in the West Health Region.

@@

clean_data %>% subset (Race=="White/Caucasian")%>%
  tbl_summary(
    percent="row",
    by= first_dose,
    include=c(Health_Region),
    sort = all_categorical() ~ "frequency"
  ) %>% bold_labels()%>%
  as_kable_extra(booktabs = TRUE,linesep="")
```

@@

>>>>

Tables A-7 and A-8 present summaries of information from the Fields survey (from the dataset analyzed in the main paper), showing responses by White/Caucasian individuals by Health Region and proportion of answers by LHIN in the West Health Region.

@@

clean_data %>% subset (Race=="White/Caucasian")%>%
  tbl_summary(
    percent="row",
    by= first_dose,
    include=c(Health_Region),
    sort = all_categorical() ~ "frequency"
  ) %>% bold_labels()%>%
  as_kable_extra(booktabs = TRUE,linesep="")

  clean_data %>%
  select(Race,Health_Region,first_dose) %>%
  tbl_strata(
    strata = Health_Region,
    .tbl_fun =
      ~ .x %>%
        tbl_summary(by = first_dose, missing = "no",percent="row") %>%
        add_n(),
    .header = "**{strata}**, N = {n}"
  )
```

@@

```{r,Central-HR-LHIN}
#| label: tbl-central-hr-lhin
#| tbl-cap: "Proportion of answers by LHIN (Central Health Region)"
#| message: false
#| echo: false

clean_data %>% subset (Health_Region=="Central")%>%
tbl_summary(include=c(LHIN),
            sort = all_categorical() ~ "frequency") %>% 
  bold_labels()%>%
  as_kable_extra(booktabs = TRUE,linesep="")

### 2023-05-12 manuscript_v5/appendix.tex
updated, working on discussion

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, interior hidden, frame hidden, breakable, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-central-hr-lhin}{}
\begin{table}
\caption{\label{tbl-central-hr-lhin}Proportion of answers by LHIN (Central Health Region) }\tabularnewline

\centering
\begin{tabular}{lc}
\toprule
\textbf{Characteristic} & \textbf{N = 1,455}\\
\midrule
\textbf{LHIN} & \\
\hspace{1em}Mississauga Halton & 495 (34\%)\\
\hspace{1em}Central & 428 (29\%)\\
\hspace{1em}Central West & 362 (25\%)\\
\hspace{1em}North Simcoe Muskoka & 170 (12\%)\\
\bottomrule
\multicolumn{2}{l}{\rule{0pt}{1em}\textsuperscript{1} n (\%)}\\
\end{tabular}
\end{table}
